Process-based phenotypes: an application to the ontology of cell physiology

### 2011-11-27 paper/drinking.tex

\author[Hoehndorf et al.]{Robert Hoehndorf$^1$ and Gabriella
  Rustici$^2$ and Midori A. Harris$^3$ and Georgios V. Gkoutos$^{1}$}

@@

\section{Introduction}

problem of phenotypes...

why reasoning, why formal, success,...

solved: structures, qualities

open: process phenotypes very complex...  
physiology...
measurements...

\subsection{Problem statement 1}
Heart rate increased: frequency of CMC increased fluid flow increased
volume per contraction increased ...

what is the relation to the heart?

\subsection{Problem statement 2}
increased apoptosis

\subsection{Observations}
observation 1: there is a fundamental distinction between attribution
of qualities for a fixed period (e.g., 24h) and single {\em act}s of
drinking.

observation 2: each of the different possible ``increased'' drinking
behavior may be the result of different underlying causes and due to
this, distinguishing between them may be important.

observation 3: depending on the experiment performed, only some of the
actual qualities can be measured.

observation 4: some qualities can be inferred given others. For
example,...

@@

\subsection{Processes, their participants and qualities}
\begin{figure}
  \centering
  \includegraphics[width=.45\textwidth]{onto.pdf}
  \caption{Overview over ...\label{fig:onto}}
\end{figure}

Most biomedical ontologies share common distinctions between different
kinds of entities: physical objects, qualities, functions and
processes. A physical object is an entity that is present as a whole
at a time point, i.e., an entity that has no temporal parts. A quality
is an attribute or feature of an entity. Physical objects together
with their qualities give rise to functions, which are capabilities or
potentials of physical objects. These functions can then be realized
by processes, which are temporally extended entities. Examples of
classes that may have processes as instances include {\em Drinking},
{\em Triathlon} or {\em Apoptosis}.  Figure \ref{fig:onto} illustrates
these basic categories of being.

Processes commonly have participants\footnote{It may be argued that
  processes necessarily have participants or that processes may have
  no participants. However, these discussions do not affect our aims
  here and take no position in this regard.}. Participants of
processes are entities that are present at time points (i.e., they
have no temporal parts) and may undergo changes during the process.
We can further distinguish different {\em roles} of participation in a
process. For example, a runner may participate in a {\em Triathlon}
process as a {\em Runner (role)}, while another person can participate
in the same process but in a different role (such as {\em Referee}).

@@

\section{Results}
\subsection{Attributes of processes}
We develop a model of process attributes that is applicable for
representations of physiology and related phenotypes. In principle, we
distinguish between three different kinds of process attributes:
process attributes that are linked directly to processes, attributes
that arise due to process parts and attributes of processes, and
attributes arising from attributes of process participants and other
process attributes.

Attribute that can be directly linked to a process are the result of
the temporal extension of a process and include process {\em duration}
and {\em temporal location}. A duration is an attribute that
characterizes the temporal extent of a process and is similar to {\em
  Length}, {\em Area} and {\em Volume} for one-, two- and
three-dimensional entities. A {\em Temporal location} positions the
time interval during which the process occurs with respect to a
reference coordinate system.

However, the majority of attributes that characterize processes are
not based on genuine process attributes alone, but rather relate
attributes of process participants with the duration of a process. In
particular, a {\em rate} typically refers to an attribute of some
entity {\em with respect to an attribute of another entity}, and in
the context of processes, rates often refer to attributes of a process
participant with respect to the duration of the process. For example,
a {\em mass flow rate} refers to the {\em Mass} of a process
participant with respect to the duration of the process, i.e., how
much matter is moved (from one point to another) through the process.
As a more complex example, a {\em rate of change of position} refers
to the {\em distance} that an object is moved with respect to the
duration of the process.

Not all rates of a process depend on attributes of the process
participants. In particular, a {\em frequency of occurrence} or {\em
  event rate} refers to the number of occurrences of a type of process
during the reference process. For example, a {\em rate of heart
  beating} refers to the number of {\em Heart beating} processes that
occur within a reference process (e.g., a process containing the heart
with a duration of one minute). Further attributes that depend on
types of processes with regard to a reference process are {\em
  distribution patterns}, i.e., how the occurrences of processes of a
particular type are distributed within a reference process. For
example, the heart may beat {\em rhythmically} or {\em arrhythmically}
within a period of time.

Related to distribution patterns are {\em changing qualities} of
processes. For example, the rate of heart beating may change ({\em
  increase} or {\em decrease}) throughout the course of the reference
process. A simple analysis of {\em increasing} ({\em decreasing})
rates would be that the rate of a heart beating within the first half
of a process if {\em lower} ({\em higher}) than in the second half of
the process. Mathematically, this process sub-division can be iterated
until we end up with infinitesimally small time intervals.

While some processes can be subdivided indefinitely without changing
the type of process, others cannot. A class $C$ of {\em continuous
  processes} is a class that has processes $p$ as instance such that
all temporal parts of $p$ are also instances of $C$. Examples include
{\em continuous movements} or {\em regulation} processes. On the other
hand, {\em discrete processes} can be subdivided into stages of
activity and stages of inactivity: a class $C$ of {\em discrete
  processes} is a class with processes $p$ as instance and there is a
part of $p$ that is not an instance of $C$. An example is {\em heart
  beating} which has periods of activity (a single heart beat) and
inactivity.

% In line with the GO, we call processes that cause

% [Regulation of X vs. X: regulation modulates the number/amount of
% occurrence...]

We may say that an object has a {\em frequency} or {\em rate} as if
these were attributes of objects at time points. For example, that my
heart {\em now} beats with a frequency of 80 beats per minute, or that
a car is moving with a speed of 280 kilometres per minute {\em at a
  particular point in time} (e.g., as observed with a speed camera)
may be considered attributes of the objects, not attributes of
processes. In our analysis, rates as attributes of objects can be
explicitly defined using rates of processes. For example, the heart
beating frequency of a particular heart $h$ at a time point $t$ is the
frequency of heart beating processes in which $h$ participates at an
infinitesimally small process around $t$ and within a reference
process. The reference process is necessary in order to obtain a value
for a frequency or rate even if the process in question is not
occurring. For example, a rate of cardiac muscle contraction can be
determined even when no cardiac muscle contraction is occurring in
virtue of a reference process with a rate of cardiac muscle
contraction.

The reference process for an object at a time point $t$ is uniquely
determined for continuous processes (the infinitesimally small process
around $t$), and ambiguous for discrete processes.

A similar construction as for continuous processes can be made for
discrete processes. A rate of blood flow is immediately an attribute
of a single heart beat process. However, during a period of time, the
cumulative rate of blood flow can be observed as the sum of the blood
flow rates of individual heart beating processes.

@@

\section{Discussion}
\subsection{Implementation as pre-composed ontology}
Pre-composed or software support for pattern?

\subsection{Use of CPO}
The Fission Yeast Phenotype Ontology (FYPO), a new ontology developed
to support annotation of phenotypes in {\em Schizosaccharomyces
  pombe}, consists of pre-composed terms describing normal or abnormal
cellular phenotypes. Over 80\% of FYPO definitions reference
descendants of GO BP 'cellular process' as the entity; a further 11\%
reference GO CC terms. All FYPO explicit definitions reference
qualities in PATO, including {\em normal}, {\em abnormal}, and several
process qualities including {\em increased duration}, {\em decreased
  occurrence}, etc. FYPO will thus fit neatly under the CPO umbrella,
and stands to benefit from the automated synchronization between CPO
and GO. {\em S. pombe} annotations to FYPO terms will provide a rich
body of highly specific, well-supported data to be integrated with
data from other species.

@@

% \subsection{Future research: acts and intentional states}
% activities/actions, intentional (goal-oriented) acts

% feeling thirsty (internal state) as driving force behind drinking
% behavior

% non-measurable, inference only

>>>>

\author[Hoehndorf et al.]{Robert Hoehndorf$^1$ and Midori
  A. Harris$^3$ and Heinrich Herre(?) and Gabriella Rustici$^2$ and
  Georgios V. Gkoutos$^{1}$}

@@

\section{Introduction}
Phenotype studies on all scales and levels of granularity are now an
invaluable tool for functional genomics research. Phenotypes of
targeted mutations in animal models are now systematically recorded to
reveal the role of individual genes within a biological system. These
phenotype studies now play a key role in translational research and
are being used to reveal candidate genes for orphan diseases and to
identify chemicals that may have effects on these diseases \cite{}.

The large volume and diversity of phenotypes within different species
and across multiple scales and levels of granularity necessitates the
application of flexible strategies for managing and integrating data
so that it becomes amenable to automated comparative analyses. To
integrate biomedical data across heterogeneous information systems,
biomedical ontologies were being developed \cite{Smith2007}. An
ontology is an explicit specification of a conceptualization of a
domain and can be used to make the meaning of terms in a vocabulary
explicit \cite{Gruber1995, Guarino1998}. They play a crucial role in
the annotation of biomedical data and the integration of model
organism databases \cite{go2010, Bada2004, goble}.

Ontologies increasingly rely on the use of Semantic Web technologies
\cite{Berners-Lee2001}. The Semantic Web provides a stack of protocols
and languages to include explicit semantics in websites. In
particular, the Web Ontology Language (OWL) \cite{Grau2008} has been
designed to express and share ontologies within the Semantic Web. OWL
is a language based on description logics (a group of formal languages
based on first-order predicate logic). Automated reasoners have been
developed within the Semantic Web to perform complex operations on
ontologies formulated in OWL. In particular, automated reasoners can
verify an ontology's consistency and use deductive inference to
perform powerful queries over ontologies. To benefit from automated
reasoning and the rapidly increasing number of software tools that are
being developed within the Semantic Web, most biomedical ontologies
are now available in OWL or can be converted into an OWL-based
representation \cite{Horrocks2007, Hoehndorf2010patterns}.

In the domain of phenotypes, multiple ontologies have been
developed. In particular, ontologies to characterize mammalian
\cite{Smith2004}, human \cite{Robinson2008}, yeast \cite{ypo} and worm
\cite{wpo} phenotypes are now available, while several more phenotype
ontologies are under development. To benefit from automated reasoning,
integrate phenotypes across species and reuse the content of anatomy
and process ontologies, classes in phenotype ontologies were defined
using the framework of the Phenotypic Attribute and Trait (PATO)
ontology \cite{Gkoutos2005}. According to the PATO framework, a
phenotype can be decomposed, using an Entity-Quality model, into an
affected entity and a quality that characterizes {\em how} the entity
is affected \cite{Gkoutos2005}. Such decompositions have been created
for several widely used phenotype ontologies \cite{Mungall2010,
  Gkoutos2009b, obml2011h1}, and are being applied together with
methods for reusing knowledge contained in anatomy ontologies
\cite{Mungall2010, Hoehndorf2010phene}.

While the PATO framework is now successfully being applied to
semantically integrate phenotypes across species, the diversity and
complexity of phenotypes in which biological processes and functions
are impaired continues to limit the interoperability between phenotype
ontologies. Major challenges for representing and integrating
process-based phenotypes include establishing the link to the
components of biological systems that have the capabilities to exhibit
such a behaviour, and that attributes of processes are often measured
{\em indirectly} and inferred from other attributes.

To illustrate these challenges, consider physiological processes of
the heart. One of the heart's functions is {\em Heart beating}, i.e.,
a capability that is realized through processes of the type {\em Heart
  beating}. {\em Blood} is a participant of {\em Heart beating}
processes.  An abnormal phenotype of an organism could be that the
{\em rate of heart beating} is increased. The intended meaning of such
a description is that the number of {\em Heart beating} processes in a
given time interval is higher than normal. Another important attribute
of heart physiology is the fluid flow rate through the heart. For an
abnormal phenotype such as {\em increased rate of fluid flow through
  the heart}, the intended meaning could either be that the amount of
fluid that is moved through the heart within a single heart beating
process is increased or that the amount of fluid that is moved through
the heart within a period of time interval is increased.

Based on these examples, we can make several observations about
process attributes. First, for a process like {\em Heart beating}, we
can distinguish between single occurrences and processes in which {\em
  Heart beating} occurs multiple times. Only the latter kind of
process may have a {\em rate of heart beating}, while {\em Heart
  beating} processes do not have such an attribute. Second, we can
distinguish between abnormal fluid flow rates in {\em Heart beating}
processes and rate of fluid flow through the heart within a given
duration. Both may have entirely different underlying causes and it is
therefore important to distinguish between them. Finally, we may be
able to infer some phenotypes from others, thereby limiting the number
of phenotypes that must be experimentally observed. For example, when
the fluid flow rate in single heart beating processes is increased and
the {\em rate of heart beating} is increased within a process $P$,
then the rate of fluid flow through the heart will be increased for
$P$.

Here, we present the foundations for an ontology of process
phenotypes. We present the outline of a theory in which several kinds
of process attributes can be distinguished so that normal and abnormal
physiology of biological systems can be formally characterized. We
apply this theory to cellular processes and create the Cell Phenotype
Ontology (CPO). The CPO is linked to reference ontologies for
qualities, biological processes, functions and cell components, and
its prime application is the unification of phenotypes on the cellular
level across different species as well as for annotation of cellular
phenotypes in domains in which no such ontology exists.

@@

\subsection{Processes and their participants}
\begin{figure}
  \centering
  \includegraphics[width=.45\textwidth]{onto.pdf}
  \caption{Overview over ...\label{fig:onto}}
\end{figure}

Most biomedical ontologies share common distinctions between different
kinds of entities: physical objects, qualities, functions and
processes. A physical object is an entity that is present as a whole
at a time point, i.e., an entity that has no temporal parts. A quality
is an attribute or feature of an entity. Physical objects together
with their qualities give rise to functions, which are capabilities or
potentials of physical objects. These functions can then be realized
by processes, which are temporally extended entities
\cite{Burek2006}. Examples of classes that may have processes as
instances include {\em Drinking}, {\em Triathlon} or {\em Apoptosis}.
Figure \ref{fig:onto} illustrates these basic categories of being.

Processes commonly have participants\footnote{It may be argued that
  processes necessarily have participants or that processes may have
  no participants. However, these discussions do not affect our aims
  here and take no position in this regard.}. Participants of
processes are entities that are present at time points (i.e., they
have no temporal parts) and may undergo changes during the process
\cite{Herre2006, Herre2010}.  We can further distinguish different
{\em roles} of participation in a process \cite{Loebe2007}. For
example, a runner may participate in a {\em Triathlon} process as a
{\em Runner (role)}, while another person can participate in the same
process but in a different role (such as {\em Referee}).

@@

\section{Results}
\subsection{Attributes of processes}
We develop a model of process attributes that is applicable for
representations of physiology and related phenotypes. In principle, we
distinguish between three different kinds of process attributes:
the first are process attributes that arise directly from processes
and include {\em duration} and {\em temporal location}; the second are
attributes that arise from processes and their parts and include {\em
  frequency}; and the third are attributes that arise from processes
and qualities of their participants, and include {\em flow rates}.

Attribute that can be directly linked to a process arise from
processes' temporal extension. For example, a duration is an attribute
that characterizes the temporal extent of a process and is similar to
{\em Length}, {\em Area} and {\em Volume} for one-, two- and
three-dimensional physical objects. A {\em Temporal location}
positions the time interval at which the process occurs with respect
to a reference coordinate system.

However, the majority of attributes that characterize processes are
not based on these types of process attributes alone, but rather
relate attributes of process participants with the duration of a
process. In particular, a {\em rate} typically refers to an attribute
of some entity {\em with respect to an attribute of another entity},
and in the context of processes, rates often refer to attributes of a
process participant with respect to the duration of the process. For
example, a {\em mass flow rate} refers to the {\em Mass} of a process
participant with respect to the duration of the process, i.e., how
much matter is moved (from one point to another) through the process.
As a more complex example, a {\em rate of change of position} refers
to the {\em distance} that an object is moved with respect to the
duration of the process.

However, not all rates of a process depend on attributes of the
process participants. In particular, a {\em frequency of occurrence}
or {\em event rate} refers to the number of occurrences of a type of
process during a reference process. For example, a {\em rate of heart
  beating} refers to the number of {\em Heart beating} processes that
occur within a reference process (e.g., a process in which the heart
participates with a duration of one minute). Further attributes that
depend on types of processes with regard to a reference process are
{\em distribution patterns}, i.e., how the occurrences of processes of
a particular type are distributed within a reference process. For
example, the heart may beat {\em rhythmically} or {\em arrhythmically}
within a period of time.

Related to distribution patterns are {\em changing qualities} of
processes. For example, the rate of heart beating may change ({\em
  increase} or {\em decrease}) throughout the course of a reference
process. A simple analysis of {\em increasing} ({\em decreasing})
rates would be that the rate of a heart beating within the first half
of a process is {\em lower} ({\em higher}) than in the second half of
the process. To make such an assertion, we divide a process into two
temporal parts. Mathematically, this process of sub-division can be
iterated until processes occur within infinitesimally small time
intervals.

While some processes can be subdivided indefinitely without changing
the type of process, others cannot. A class $C$ of {\em continuous
  processes} is a class that has processes $p$ as instance such that
all temporal parts of $p$ are also instances of $C$. Examples include
{\em continuous movements} or {\em mass flow} processes. On the other
hand, {\em discrete processes} can be subdivided into stages of
activity and stages of inactivity: a class $C$ of {\em discrete
  processes} is a class with processes $p$ as instance and there is a
part of $p$ that is not an instance of $C$. An example is {\em heart
  beating} which has periods of activity (a single heart beat) and
inactivity.

We may further attribute a {\em frequency} or {\em rate} to an object
instead of a process. For example, a heart that beats {\em now} with a
frequency of 80 beats per minute, or a car that is moving at a speed
of 180 kilometres per hour {\em at a particular point in time} (e.g.,
as observed with a speed camera) can be considered attributes of the
objects (the heart or the car), not attributes of the processes in
which the objects participate. However, these are {\em different}
kinds of attributes. Rates, when considered as attributes of objects,
may be explicitly defined using rates of processes. For example, the
heart beating frequency of a particular heart $h$ at a time point $t$
is the frequency of a reference heart beating process in which $h$
participates. Such a reference process is necessary in order to obtain
a value for a frequency even when no {\em heart beating} process is
occurring. However, the frequency is only an attribute of the heart in
virtue of such a reference process in which {\em heart beating} is
actually occurring.  This reference process can be uniquely determined
for continuous processes (such that the frequency or rate of an object
at a time $t$ is the frequency or rate of the infinitesimally small
process around $t$), and is ambiguous for discrete processes (in which
case the reference process must be explicitly stated).

Figure \ref{fig:patterns} illustrates some examples of
non-comparative and comparative process attributes.

% A similar construction as for continuous processes can be made for
% discrete processes. A rate of blood flow is immediately an attribute
% of a single heart beat process. However, during a period of time, the
% cumulative rate of blood flow can be observed as the sum of the blood
% flow rates of individual heart beating processes.

@@

\section{Discussion}
\subsection{Use of CPO}
The Fission Yeast Phenotype Ontology (FYPO), a new ontology developed
to support annotation of phenotypes in {\em Schizosaccharomyces
  pombe}, consists of pre-composed terms describing normal or abnormal
cellular phenotypes. Over 80\% of FYPO definitions reference
descendants of GO BP 'cellular process' as the entity; a further 11\%
reference GO CC terms. All FYPO explicit definitions reference
qualities in PATO, including {\em normal}, {\em abnormal}, and several
process qualities including {\em increased duration}, {\em decreased
  occurrence}, etc. FYPO will thus fit neatly under the CPO umbrella,
and stands to benefit from the automated synchronization between CPO
and GO. {\em S. pombe} annotations to FYPO terms will provide a rich
body of highly specific, well-supported data to be integrated with
data from other species.

@@

\subsection{Pre- vs. post-composed ontology}
Pre-composed or software support for pattern?
